Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-15 analysis/paper/article.qmd
attempt to standardize captilisation of topics and methods throughout

Landscape archaeology, although dominant, has fluctuated over the years and has been on a downtrend since 2018-2019. Computational and digital archaeology is the second most represented method, showing an overall increasing trend, particularly since 2020-2021. Archaeobotany shows a strong cyclical trend, rising, falling, and then rising again over our study period. Archaeometry and Geoarchaeology have maintained a relatively low but steady presence in job ads, peaking in 2017-2018 and 2018-2019 and declining thereafter. Lithic analysis and Zooarchaeology are also mentioned relatively infrequently in job ads and show an inverse correlation with each other after 2018-2019.

@@

The most frequently mentioned topic in this sample of job ads is Environmental archaeology [@fig-show-topi-trends]. This category encompasses such phrases as human-environmental dynamics, interaction between humans and their environments, environmental change, climate change, historical ecology, ecological knowledge, human ecology, and ecological systems. Public archaeology is the second most frequent topic overall; this category included phrases such as cultural resource management, cultural heritage, heritage studies, museum studies, human rights, community engaged, historic preservation, social justice, community-based, repatriation, and community-engaged archaeology. The least common topics in our sample are Pleistocene archaeology (e.g. human origins, hunter-gatherer archaeology) and digital archaeology.

The years 2019-2020 and 2020-2021 show striking changes in the popularity of topics in job ads. Indigenous and historical archaeology (which includes archaeology of the African diaspora and enslaved people) became the most popular topic at this time, rising from being one of the least popular topics from 2012-2017. Conversely, archaeological science, which was popular during 2012-2017, was rarely mentioned in job ads during 2019-2021. Ancient Europe and the Mediterranean, an infrequently mentioned topic for the entire study period, virtually disappeared from job ads during 2019-2021. Mentions of complex societies in ads decrease at a steady rate, while mentions of public archaeology consistently increase over the study period. The topic of Environmental Archaeology remains high over time.

@@

In our sample, job ads were more topically rich than geographically or methodologically rich. That is, ads were more likely to mention multiple topics than they were to mention multiple methods or geographic locations. A Kruskal-Wallis test indicated significantly higher richness in topics compared to richness of geographic locations or methods in job ads (χ2 (df = `r unname(kw_test_richness$parameter)`, N = `r nrow(kw_test_richness_input)`) = `r round(unname(kw_test_richness$statistic), 2)`, p = `r expSup(unname(kw_test_richness$p.value))`). @fig-show-cooc shows topic co-occurrences in our sample. Indigenous and historical archaeology often occurs in job ads with Public archaeology and North American archaeology. Complex societies and environmental archaeology were frequently found in the same ads. Bioarchaeology, Archaeological science, and Evolutionary archaeology are another cluster of topics that frequently co-occur. Other topics are relatively isolated. For example, Pleistocene archaeology and digital archaeology rarely occur with other topics.

@@

Overall, we found more tenure-track jobs advertised each year than non-tenure-track, with the exception of 2013-2014. The wiki’s emphasis on tenure-track jobs runs counter to well-established patterns in the academic job market, wherein short-term and contingent positions are fast outpacing “the last good job[s] in America” [@aronowitz2001]. Research across anthropological sub-fields has shown that the number of tenure-track positions advertised within the discipline has also declined. Analyzing the job ads posted in *Anthropology News* in 1999–20000 versus 2019–2020, for example, Gershon and Rachok [-@HelloTristesTropes2021] found that the number of tenure-track positions decreased by 25% between their two time slices. Similarly, in his examination of jobs ads posted on the Biological Anthropology Academic Job Wiki between 2010–2017, Passalacqua noted that approximately one quarter of the 474 ads posted were for adjunct or visiting staff [@Passalacqua_2018, pp.773]. Though Passalacqua did not examine how the proportion of job types changed over time, he did underscore “a current diverging trend of decreasing academic job advertisements and increasing doctoral degrees in biological anthropology which could result in serious consequences for the discipline” [-@Passalacqua_2018, pp.773]. Such observations have a surprising longevity. Rogge raised the alarm nearly half a century ago, warning of the “maladaptive” consequences of the exponential rate of disciplinary growth and impending consequences for job seekers [@rogge1976, pp.839]. The dwindling number of permanent academic positions is not a trend unique to anthropology—as of 2019, the AAUP documented a 36% increase in contingent positions over the preceding 15 years (AAUP 2022), and as of 2022, contingent positions make up more than half of faculty positions in the United States [@colby2022annual] .

@@

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as the Archaeology of complex societies and Archaeological science, the geographic foci of Mesoamerica and South America, and methods relating to Landscape archaeology <!--There is some shifting throughout the paper as to whether we capitalize the first letter of region/methods/topics or not. I think we should be consistent and capitalize throughout-->. These trends are harder to explain as we cannot link their origins to a historical event such as the COVID-19 pandemic. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives, in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological; @Fogelin2019] <!--Should we cite Cate Frieman here? BM: grimace emoji -->. This theoretical trend might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past and an increased concern with climate change, environmental sustainability, and resilience, shifting attention away from the study of monumental architecture, elite societies, political hierarchies, and state systems.

>>>>

Landscape archaeology, although dominant, has fluctuated over the years and has been on a downtrend since 2018-2019. Computational and Digital archaeology is the second most represented method, showing an overall increasing trend, particularly since 2020-2021. Archaeobotany shows a strong cyclical trend, rising, falling, and then rising again over our study period. Archaeometry and Geoarchaeology have maintained a relatively low but steady presence in job ads, peaking in 2017-2018 and 2018-2019 and declining thereafter. Lithic analysis and Zooarchaeology are also mentioned relatively infrequently in job ads and show an inverse correlation with each other after 2018-2019.

@@

The most frequently mentioned topic in this sample of job ads is Environmental archaeology [@fig-show-topi-trends]. This category encompasses such phrases as human-environmental dynamics, interaction between humans and their environments, environmental change, climate change, historical ecology, ecological knowledge, human ecology, and ecological systems. Public archaeology is the second most frequent topic overall; this category included phrases such as cultural resource management, cultural heritage, heritage studies, museum studies, human rights, community engaged, historic preservation, social justice, community-based, repatriation, and community-engaged archaeology. The least common topics in our sample are Pleistocene archaeology (e.g. human origins, hunter-gatherer archaeology) and Digital archaeology.

The years 2019-2020 and 2020-2021 show striking changes in the popularity of topics in job ads. Indigenous and historical archaeology (which includes archaeology of the African diaspora and enslaved people) became the most popular topic at this time, rising from being one of the least popular topics from 2012-2017. Conversely, archaeological science, which was popular during 2012-2017, was rarely mentioned in job ads during 2019-2021. Ancient Europe and the Mediterranean, an infrequently mentioned topic for the entire study period, virtually disappeared from job ads during 2019-2021. Mentions of Complex societies in ads decrease at a steady rate, while mentions of public archaeology consistently increase over the study period. The topic of Environmental Archaeology remains high over time.

@@

In our sample, job ads were more topically rich than geographically or methodologically rich. That is, ads were more likely to mention multiple topics than they were to mention multiple methods or geographic locations. A Kruskal-Wallis test indicated significantly higher richness in topics compared to richness of geographic locations or methods in job ads (χ2 (df = `r unname(kw_test_richness$parameter)`, N = `r nrow(kw_test_richness_input)`) = `r round(unname(kw_test_richness$statistic), 2)`, p = `r expSup(unname(kw_test_richness$p.value))`). @fig-show-cooc shows topic co-occurrences in our sample. Indigenous and historical archaeology often occurs in job ads with Public archaeology and North American archaeology. Complex societies and Environmental archaeology were frequently found in the same ads. Bioarchaeology, Archaeological science, and Evolutionary archaeology are another cluster of topics that frequently co-occur. Other topics are relatively isolated. For example, Pleistocene archaeology and Digital archaeology rarely occur with other topics.

@@

Overall, we found more tenure-track jobs advertised each year than non-tenure-track, with the exception of 2013-2014. The wiki’s emphasis on tenure-track jobs runs counter to well-established patterns in the academic job market, wherein short-term and contingent positions are fast outpacing "the last good job[s] in America" [@aronowitz2001]. Research across anthropological sub-fields has shown that the number of tenure-track positions advertised within the discipline has also declined. Analyzing the job ads posted in *Anthropology News* in 1999–20000 versus 2019–2020, for example, Gershon and Rachok [-@HelloTristesTropes2021] found that the number of tenure-track positions decreased by 25% between their two time slices. Similarly, in his examination of jobs ads posted on the Biological Anthropology Academic Job Wiki between 2010–2017, Passalacqua noted that approximately one quarter of the 474 ads posted were for adjunct or visiting staff [@Passalacqua_2018, pp.773]. Though Passalacqua did not examine how the proportion of job types changed over time, he did underscore "a current diverging trend of decreasing academic job advertisements and increasing doctoral degrees in biological anthropology which could result in serious consequences for the discipline" [-@Passalacqua_2018, pp.773]. Such observations have a surprising longevity. Rogge raised the alarm nearly half a century ago, warning of the "maladaptive" consequences of the exponential rate of disciplinary growth and impending consequences for job seekers [@rogge1976, pp.839]. The dwindling number of permanent academic positions is not a trend unique to anthropology—as of 2019, the AAUP documented a 36% increase in contingent positions over the preceding 15 years (AAUP 2022), and as of 2022, contingent positions make up more than half of faculty positions in the United States [@colby2022annual] .

@@

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as Complex societies and Archaeological science, the geographic foci of Mesoamerica and South America, and methods relating to Landscape archaeology. These trends are harder to explain as we cannot link their origins to a historical event such as the COVID-19 pandemic. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives, in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological; @Fogelin2019] <!--Should we cite Cate Frieman here? BM: grimace emoji -->. This theoretical trend might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past and an increased concern with climate change, environmental sustainability, and resilience, shifting attention away from the study of monumental architecture, elite societies, political hierarchies, and state systems.
